Type theories as quotient inductive-recursive types

### 2025-02-17 tex/main.tex
move many proofs to the appendix

To ensure that type substitution is strict for all %\emph
{rules}, we turn to a definition by overlapping patterns (\cref{sec:meta:overlapping}) instead to make each clause strict. 
In addition, we can include other derived rules such as $[\pi_1(\sigma, t)]_{\Ty}\;A = [ \sigma]_{\Ty}\;A$ as long as they are locally confluent and terminating.
We end up with the following definition for type substitution:
\begin{alignat}{3}
[\blank]_{\Ty} \blank            &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \nonumber \\
[ \idS ]_{\Ty}\;A                & \reduce A \label{eq:def-type-subst-1}\\
[ \sigma ; \tau ]_{\Ty}\;A       & \reduce [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:def-type-subst-2}\\
[ \pi_1(\sigma, t) ]_{\Ty}\;A    & \reduce [\sigma]_\Ty\;A                       \label{eq:def-type-subst-3} \\
[ \pi_1(\sigma; \tau) ]_{\Ty}\;A & \reduce [\sigma]_\Ty\;([\pi_1\tau]_\Ty\;A)    \label{eq:def-type-subst-4} \\
[ \sigma ]_{\Ty}\;\UU            & \reduce \UU                                   \label{eq:def-type-subst-5}
\end{alignat}
We first check that this definition of type substitution is confluent.
\begin{proposition}[Termination]
  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
  \danger
\end{proposition}
%\begin{proof}
%  Clearly terms on the RHS are structurally smaller~\cite{Abel2002}.
%\end{proof}
\begin{proposition}[Local confluence] \label{prop:local-confluence-1}
  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
  \danger
\end{proposition}
\begin{proof}
  To show the local confluence, we consider all local peaks, i.e.\ each combination of \eqref{eq:def-type-subst-5} with \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4},
  \[
    [\idS]_{\Ty}\;\UU, \qquad [\sigma;\tau]_{\Ty}\;\UU, \qquad [\pi_1(\sigma, t)]_{\Ty}\;\UU, \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;\UU
  \]
  and show that each applicable clause does result in $\UU$.

@@

As we are defining a function on a quotient inductive type, we have to prove that $[\blank]_{\Ty}\blank$ is coherent, i.e., that it respects all equality constructors.
%\LT[noinline]{Should we mention any general condition for coherence?}
%\Fredrik[noinline]{Is there one?}
%\LT[noinline]{I think Cubical Agda has sort of conditions? (Cubical Agda does not support rewriting)}
\begin{proposition}[Coherence]\label{prop:coherence-1}
  $[\sigma]_{\Ty}\,A$ is (propositionally) equal to $[\tau]_{\Ty}\,A$ for every equality constructor $p : \sigma = \tau$.
  \danger
\end{proposition}
\begin{proof}
  As equality constructors $[\idS]_T$, $[;]_T$, and $[]\UU$ for types become strict, it suffices to show that the following identities
  \begin{align*}
    [\sigma;\idS]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
                                     & [\idS;\sigma]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
                                     & [(\sigma;\tau);\gamma]_{\Ty} \;A & = [\sigma; (\tau; \gamma)]_{\Ty}A, \\
    [\pi_1(\sigma, t)]_{\Ty}\;A      & = [\sigma]_{\Ty}\;A, \\
    [\sigma; (\tau, t)]_{\Ty} \;A    & = [(\sigma;\tau), [\sigma]\, t]_{\Ty}\;A,
                                     & [\sigma]_{\Ty} \;A               & = [\emptyctx]_{\Ty}\;A,
                                     & [\sigma]_{\Ty} \;A               & = [\pi_1\sigma, \pi_2\sigma]_{\Ty}\;A
  \end{align*}
  hold propositionally.
  The first two identities of the first row hold strictly by \eqref{eq:def-type-subst-1} and \eqref{eq:def-type-subst-2}, while terms of the both sides of the third reduce to $[\sigma]_{\Ty}\;([\tau]_{\Ty}\;([\gamma]_{\Ty}\;A))$ by \eqref{eq:def-type-subst-2}.

@@

Then, the equality constructor $[]\El$ is accordingly changed to 
\begin{align}
  [ \sigma ]_{\Ty}\,(\El\, u) & \reduce \El\,([\sigma]_{\Tm} {u}) \label{eq:def-type-subst-7}
\end{align}
to retain the local confluence for type substitution:
\begin{proposition}[Termination]
  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
  \danger
\end{proposition}
%\begin{proof}
%  Again, each recursive call is on a structurally smaller term~\cite{Abel2002}.
%\end{proof}
\begin{proposition}[Local confluence] \label{prop:local-confluence-2}
  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
  \danger
\end{proposition}
\begin{proof}
  In addition to previous cases in the proof of \cref{prop:local-confluence-1}, consider terms 
  \[
    [\idS]_{\Ty}\;(\El\,u), \qquad [\sigma;\tau]_{\Ty}\;(\El\,u), \qquad [\pi_1(\sigma, t)]_{\Ty}\;(\El\,u), \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;(\El\,u)
  \]
  which reduce to the following terms
  \[
    \El\,u, \qquad \El([\sigma]_{\Tm}\,[\tau]_{\Tm}\,u), \qquad \El([\sigma]_{\Tm}\,u), \qquad\text{and}\qquad
    \El([\sigma]_{\Tm}\,[\pi_1\tau]_{\Tm}\,u)
  \]
  respectively, no matter which function clause of $[\blank]_{\Ty}\blank$ is used.
  Therefore, type substitution is locally confluent.
\end{proof}

@@

\begin{proposition}\label{prop:correctness-1}
  For any $\sigma : \Sub\;\Gamma\;\Delta$, $A : \Ty\;\Delta$, and $t : \Tm\;\Delta\;A$, the term $[\sigma]t$ is propositionally equal to $[\sigma]_{\Tm}\,t$.
\end{proposition}
\begin{proof}
  We show this statement by induction on $\sigma$. Since we are proving a propositional equality, and we are assuming UIP, all equality constructors are automatically respected, so we only need to worry about the point constructors.
  \begin{enumerate}
    \item For the identity substitution $\idS$, we have $[\idS]t = t \judgeq [\idS]_{\Tm} t$ by definition.
    \item For a composite substitution $\sigma; \tau$, we have
      \begin{align*}
        [\sigma;\tau]t & = [\sigma]\,([\tau] t)            && \text{by $[;]t$} \\
%                       & = [\sigma]\,([\tau]_{\Tm}\,t)       && \text{by the induction hypothesis $[\tau]t = [\tau]_{\Tm}\,t$ for any $t$} \\
                       & = [\sigma]_{\Tm}\,([\tau]_{\Tm}\,t) && \text{by the induction hypothesis (twice)} \\ %$[\sigma]u = [\sigma]_{\Tm}\,u$ for any $u$} \\
                       & \judgeq [\sigma;\tau]_{\Tm}\,t && \text{by definition.}
      \end{align*}
    \item For $\pi_1(\sigma, t)$, we have
      \begin{align*}
        [\pi_1(\sigma, t)]t & = [\sigma] t               && \text{by $\pi_1\beta$} \\
                            & = [\sigma]_{\Tm} \,t       && \text{by the induction hypothesis} \\ % $[\tau]t = [\tau]_{\Tm}\,t$ for any $t$} \\
                            & \judgeq [\pi_1(\sigma, t)]_{\Tm}\,t && \text{by definition.}
      \end{align*}
    %\item For $\pi_1(\sigma; \tau)$, it is proved similarly as the previous case.
    \item For the remaining cases, $[\sigma]_{\Tm} t$ is equal to $[\sigma]\,t$ strictly.
  \end{enumerate}
  Hence, we conclude that $[\blank]_{\Tm}\blank$ is propositionally equal to $[\blank]\blank$.
\end{proof}

Finally, we have to show that both $[\blank]_{\Ty}\blank$ and $[\blank]_{\Tm}\blank$ are coherent.
\begin{proposition}[Coherence] \label{prop:coherence-2}
  For every equality constructor $p : \sigma = \tau : \Sub\;\Gamma\;\Delta$, the following identities
  \danger
  \[
    [\sigma]_{\Ty}\,A = [\tau]_{\Ty}\,A
    \quad\text{and}\quad
    [\sigma]_{\Tm}\,t = [\tau]_{\Tm}\,t
  \]
  hold for any type $A$ and any term $t$.
\end{proposition}
\begin{proof}
  By \cref{re:coherence-proof}, for type substitution, we only have to consider the following cases 
  \begin{align*}
    [\sigma; (\tau, t)]_{\Ty} \;A    & = [(\sigma;\tau), [\sigma]\, t]_{\Ty}\;A,
                                     & [\sigma]_{\Ty} \;A               & = [\emptyctx]_{\Ty}\;A,
                                     & [\sigma]_{\Ty} \;A               & = [\pi_1\sigma, \pi_2\sigma]_{\Ty}\;A
  \end{align*}
  while other cases hold strictly.
  Again, for each case, we prove them by induction on $A$ and consider the case $A \defeq \El\,u$ only which is added to the definition of $[\blank]_{\Ty}\blank$ in this section. 
  However, all cases follow easily from \cref{prop:correctness-1}.
  For example, consider the first case:
  \begin{align*}
    [\sigma; (\tau, t)]_{\Ty}(\El\,u) & \judgeq \El([\sigma; (\tau, t)]_{\Tm}\,u) && \text{by \eqref{eq:def-type-subst-7}} \\
                                      & = \El( [\sigma; (\tau, t)]\,u)            && \text{by \cref{prop:correctness-1}} \\
                                      & = \El( [\sigma; \tau, [\sigma]t)]\,u)     && \text{by $\mathsf{concat}$} \\
                                      & = \El( [\sigma; \tau, [\sigma]t)]_{\Tm}\,u) && \text{by \cref{prop:correctness-1}} \\
                                      & \judgeq [\sigma; \tau, [\sigma]t)]_{\Ty}\left(\El\,u\right) && \text{by \eqref{eq:def-type-subst-7} }
  \end{align*}
  The remaining two cases are omitted.

@@

\begin{proposition}[Termination]
  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
  \danger
\end{proposition}
\begin{proposition}[Local confluence]\label{prop:local-confluence-3}
  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
  \danger
\end{proposition}
\begin{proof}
  In addition to previous cases in \cref{prop:local-confluence-1,prop:local-confluence-2}, consider
  \begin{align*}
    [\idS]_{\Ty}\;(\Pi\;A\;B) && [\sigma;\tau]_{\Ty}\;(\Pi\;A\;B) && [\pi_1(\sigma, t)]_{\Ty}\;(\Pi\;A\;B) && [\pi_1(\sigma; \tau)]_{\Ty}\;(\Pi\;A\;B).
  \end{align*}
  Each of them reduces to the following terms after applying clauses of $\blank\uparrow\blank$ and $[\blank]_{\Ty}\blank$:
  \begin{align*}
    & \Pi\;A\;B                && \Pi\;([\sigma]_{\Ty}\,[\tau]_{\Ty}\,A)\;([\sigma \uparrow [\tau]\,A ]_{\Ty}\,[\tau \uparrow A ]_{\Ty}\,B) \\
    & \Pi\;([\sigma]_{\Ty}\;A)([\sigma \uparrow A]_{\Ty}\;B) && \Pi\;([\sigma]_{\Ty}\,[\pi_1\tau]_{\Ty}\,A)\;([\sigma \uparrow [\pi_1\tau]_{\Ty}\,[\pi_1\tau \uparrow A]_{\Ty}\,B)
  \end{align*}
  respectively.
  Hence, the type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
\end{proof}

@@

\begin{proposition} \label{prop:correctness-2}
  For every $\sigma : \Sub\;\Gamma\;\Delta$, and $A : \Ty\;\Delta$, the lifting
  $\sigma \uparrow A$ is propositionally equal to $\sigma^+$. 
\end{proposition}
\begin{proof}
  We prove the statement by induction on $\sigma$ with \cref{lem:lifting}.
  Every case follows from equational reasoning straightforwardly.
  For example, if $\sigma$ is a composite $\sigma;\tau$, we have
  \begin{align*}
    (\sigma;\tau) \uparrow A & \judgeq (\sigma \uparrow [\tau]\,A);(\tau \uparrow A) && \text{by definition} \\
                             & = \sigma^+;\tau^+                                     && \text{by induction hypothesis} \\ 
                             & = (\sigma;\tau)^+                                     && \text{by \cref{lem:lifting}.} 
  \end{align*}
  Other cases follow similarly.
\end{proof}

@@

Finally, we show that these simultaneously defined functions are coherent.
\begin{proposition}[Coherence]
  For every constructor $p : \sigma = \tau : \Sub\;\Gamma\;\Delta$, we have
  \danger
  \[
    \sigma \uparrow A =^{\lam{\gamma}\Sub\;(\Gamma, [ \gamma ]_{\Ty} A)\;(\Delta, A)}_{\alert{p}} \tau \uparrow A
    \qquad\text{and}\qquad
    [\sigma]_{\Tm}\,t =^{\lam{\gamma}\Tm\;\Gamma\;([\gamma]_{\Ty}A)}_{\alert{p}} [\tau]_{\Tm}\,t,
  \]
  for any type $A$ and any term $t$.
\end{proposition}
\begin{proof}
  These two identities follow from \cref{prop:correctness-2,prop:correctness-3} easily with the heterogeneous identity (see \cref{sec:meta}).
\end{proof}

@@

Of course, we also have to prove coherence for the newly introduced recursive functions to verify that they are well defined.
\begin{proposition}[{Coherence for $[\blank]_{\Tel}$ and $\blank\upuparrows\blank$}]
  \danger
  For any equality constructor $p : \sigma = \tau$, we have
  $[\sigma]_{\Tel}\,\Xi = [\tau]_{\Tel}\,\Xi$ and $\sigma \upuparrows \Xi =^{\Sub\;(\Gamma\dplus [\sigma]_{\Tel}\Xi)\;(\Delta\dplus\Xi)} \tau \upuparrows \Xi$
  for any telescope $\Xi$.
\end{proposition}

@@

\begin{corollary}[Coherence]
  The identity $[\sigma]_{\Ty}\,A = [\tau]_{\Ty}\,A$ holds for every equality constructor $p : \sigma = \tau$.
  \danger
\end{corollary}

@@

The elimination principle computes by interpreting each constructor using the corresponding method inductively, e.g.,
\begin{alignat*}{5}
%  \elim_{\Ctx}\;\emptyctx   & \reduce \emptyctx_{\Ctx}                                    && : \Ctx^M\;\emptyctx  \\
  \elim_{\Ctx}\;(\Gamma, A) & \reduce (\elim_{\Ctx}\,\Gamma),_{\Ctx}^{M} (\elim_{\Ty}\,A) && : \Ctx^M\;(\Gamma, A) %\\
%  \elim_{\Ty}\;\UU          & \reduce \UU^M && :\Ty^M\;\Gamma^M\;\UU \\
%  \elim_{\Sub}\;\emptysub   & \reduce \emptysub^M_{\Sub} &&: \Sub^M\;\Gamma^M\;\emptyctx^M_{\Ctx}\;\emptyctx \\
%                            & \vdots
\end{alignat*}
%
Interestingly, a key difference compared to its QIIT counterpart is that $[\blank]_{\Ty}\blank$, $[\blank]_{\Tm}$, and $\blank\uparrow\blank$ are functions rather than (formed by) constructors, meaning they cannot be eliminated inductively but instead become propositional identities.
%type substitution $[\blank]\blank$ was a constructor and $[\sigma]\,A$ was eliminated to $[\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A)$; similarly for term substitution.
\begin{proposition}\label{prop:provable-function-clauses}
  Given the motives and the methods, we can prove
\begin{alignat*}{5}
  \elim_{\Ty}[] & : \elim_{\Ty}\,([\sigma]_{\Ty}\,A) && = && [\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A) \\
  \elim_{\Tm}[] & : \elim_{\Tm}\,([\sigma]_{\Tm}\,t) && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{\elim_{\Ty}[]}} && [\elim_{\Sub}\,\sigma]^M_{\Tm}\,(\elim_{\Tm}\,t) \\
  \elim_{\Sub}[] & : \elim_{\Sub}(\sigma\uparrow A) && =^{\lam{B^M}\Sub^M\;(\Gamma^M,^M B^M)\;(\Delta^M, A^M)\;(\sigma \uparrow A)}_{\alert{\elim_{\Ty}[]}} && (\elim_{\Sub} \sigma) \uparrow^M (\elim_{\Ty}\,A) 
\end{alignat*}
\end{proposition}

@@

Going back and forth from one definition to the other gives us the equivalences
\begin{theorem}
  Type theory defined as a QIIRT and a QIIT are equivalent, i.e.\ 
  \begin{itemize}
    \item $\eqv{\Ctx_{\IR}}{\Ctx_{\I}}$, 
    \item $\eqv{\left((\Gamma : \Ctx_{\IR})\to\Ty_{\IR}\;\Gamma\right)}{\left((\Gamma : \Ctx_{\I})\to\Ty_{\I}\;\Gamma\right)}$,
    \item $\eqv{\left((\Gamma\;\Delta : \Ctx_{\IR}) \to \Sub_{\IR}\;\Gamma\;\Delta\right)}%
               {\left((\Gamma\;\Delta: \Ctx_{\I}) \to \Sub_{\I}\;\Gamma\;\Delta\right)}$, and 
    \item $\eqv{\left((\Gamma : \Ctx_{\IR}) \to (A : \Ty_{\IR}\;\Gamma) \to \Tm_{\IR}\;\Gamma\;A\right)}
               {\left((\Gamma : \Ctx_{\I}) \to (A : \Ty_{\I}\;\Gamma) \to \Tm_{\I}\;\Gamma\;A\right)}$.
  \end{itemize}
\end{theorem}
%\Fredrik{Why only on contexts? If that's the limit of the proof at the moment, I think it would be better to state just the translations, and maybe have a running text remark after the theorem.}

@@

\section{Omitted proofs}

\Fredrik{Do properly with restate-thm etc}
The first set of equality constructors are easy to prove, since the lifting and substitution operations compute independently for these cases:
\begin{proposition}[Coherence, I]
  \danger
  For the following equality constructors $p : \sigma = \tau$
  \[
    \mathsf{idr} : \sigma;\idS = \sigma, \quad
    \mathsf{idl} : \idS;\sigma = \sigma, \quad\text{and}\quad
    \textnormal{;-}\mathsf{assoc} : (\sigma;\tau);\gamma = \sigma; (\tau; \gamma), 
  \]
  the identities $[\sigma]_{\Tel}\,\Xi \judgeq [\tau]_{\Tel}\,\Xi$ and $[\sigma ]_{\Ty}\,A \judgeq [\tau ]_{\Ty}\,A$
  hold strictly for any telescope $\Xi$ and any type $A$ and
  \[
    \sigma \upuparrows \Xi =^{\Sub\;(\Gamma\dplus [\sigma]_{\Tel}\Xi)\;(\Delta\dplus\Xi)}
    \tau   \upuparrows \Xi
  \]
  holds propositionally for any telescope $\Xi$.
\end{proposition}
\begin{proof}
  Thanks to overlapping patterns, the coherence of $[\blank]_{\Tel}\blank$ and $[\blank]_{\Ty}\blank$ for these cases holds strictly by definition and the coherence of $\blank \upuparrows \blank$ boils down to the equality constructor itself.
  Take $\mathsf{idr}$ as an example, we can see 
  \begin{align*}
    (\sigma; \idS) \upuparrows \Xi & \judgeq (\sigma \upuparrows [\idS]_{\Tel} \Xi) ; (\idS \upuparrows \Xi) \\
                                   & \judgeq (\sigma \upuparrows \Xi) ; \idS \\
                                   & = \sigma \upuparrows \Xi
  \end{align*}
  where the last identity follows from $\mathsf{idr}:(\sigma \upuparrows \Xi) ; \idS = \sigma \upuparrows \Xi$.

>>>>

To ensure that type substitution is strict for all %\emph
{rules}, we turn to a definition by overlapping patterns (\cref{sec:meta:overlapping}) instead to make each clause strict. 
In addition, we can include other derived rules such as $[\pi_1(\sigma, t)]_{\Ty}\;A = [ \sigma]_{\Ty}\;A$ as long as they are locally confluent and terminating.
We end up with the following definition for type substitution:
\begin{alignat}{3}
[\blank]_{\Ty} \blank            &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \nonumber \\
[ \idS ]_{\Ty}\;A                & \reduce A \label{eq:def-type-subst-1}\\
[ \sigma ; \tau ]_{\Ty}\;A       & \reduce [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:def-type-subst-2}\\
[ \pi_1(\sigma, t) ]_{\Ty}\;A    & \reduce [\sigma]_\Ty\;A                       \label{eq:def-type-subst-3} \\
[ \pi_1(\sigma; \tau) ]_{\Ty}\;A & \reduce [\sigma]_\Ty\;([\pi_1\tau]_\Ty\;A)    \label{eq:def-type-subst-4} \\
[ \sigma ]_{\Ty}\;\UU            & \reduce \UU                                   \label{eq:def-type-subst-5}
\end{alignat}
We first check that this definition of type substitution is confluent.
%
\begin{proposition}\label{prop:local-confluence-1}
  Type substitution $[\blank]_{\Ty}\blank$ is terminating and locally confluent.
  \danger
\end{proposition}
\begin{proof}
  Termination is clear, since all recursive calls are on structurally smaller arguments~\cite{Abel2002}.
  To show the local confluence, we consider all local peaks, i.e.\ each combination of \eqref{eq:def-type-subst-5} with \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4},
  \[
    [\idS]_{\Ty}\;\UU, \qquad [\sigma;\tau]_{\Ty}\;\UU, \qquad [\pi_1(\sigma, t)]_{\Ty}\;\UU, \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;\UU
  \]
  and show that each applicable clause does result in $\UU$.

@@

As we are defining a function on a quotient inductive type, we have to prove that $[\blank]_{\Ty}\blank$ is coherent, i.e., that it respects all equality constructors.
%\LT[noinline]{Should we mention any general condition for coherence?}
%\Fredrik[noinline]{Is there one?}
%\LT[noinline]{I think Cubical Agda has sort of conditions? (Cubical Agda does not support rewriting)}
\begin{proposition}\label{prop:coherence-1}
  For every equality constructor $p : \sigma = \tau$, we have $[\sigma]_{\Ty}\,A = [\tau]_{\Ty}\,A$.
  \danger
\end{proposition}
\begin{proof}
  As equality constructors $[\idS]_T$, $[;]_T$, and $[]\UU$ for types become strict, it suffices to show that the following identities
  \begin{align*}
    [\sigma;\idS]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
                                     & [\idS;\sigma]_{\Ty} \;A          & = [\sigma]_{\Ty}\;A,
                                     & [(\sigma;\tau);\gamma]_{\Ty} \;A & = [\sigma; (\tau; \gamma)]_{\Ty}A, \\
    [\pi_1(\sigma, t)]_{\Ty}\;A      & = [\sigma]_{\Ty}\;A, \\
    [\sigma; (\tau, t)]_{\Ty} \;A    & = [(\sigma;\tau), [\sigma]\, t]_{\Ty}\;A,
                                     & [\sigma]_{\Ty} \;A               & = [\emptyctx]_{\Ty}\;A,
                                     & [\sigma]_{\Ty} \;A               & = [\pi_1\sigma, \pi_2\sigma]_{\Ty}\;A
  \end{align*}
  hold propositionally.
  The first two identities of the first row hold strictly by \eqref{eq:def-type-subst-1} and \eqref{eq:def-type-subst-2}, while terms of the both sides of the third reduce to $[\sigma]_{\Ty}\;([\tau]_{\Ty}\;([\gamma]_{\Ty}\;A))$ by \eqref{eq:def-type-subst-2}.

@@

Then, the equality constructor $[]\El$ is accordingly changed to 
\begin{align}
  [ \sigma ]_{\Ty}\,(\El\, u) & \reduce \El\,([\sigma]_{\Tm} {u}) \label{eq:def-type-subst-7}
\end{align}
to retain the local confluence for type substitution. The proof is similar to the proof of \cref{prop:local-confluence-1}, and can be found in \cref{appendix:proofs}.
%\begin{proposition}[Termination]
%  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
%  \danger
%\end{proposition}
%\begin{proof}
%  Again, each recursive call is on a structurally smaller term~\cite{Abel2002}.
%\end{proof}
\begin{restatable}{proposition}{propConfluenceTwo}\label{prop:local-confluence-2}
  Type substitution $[\blank]_{\Ty}\blank$ is terminating and locally confluent.
  \qed
  \danger
\end{restatable}

@@

\begin{restatable}{proposition}{propCorrectness}\label{prop:correctness-1}
  For any $\sigma : \Sub\;\Gamma\;\Delta$, $A : \Ty\;\Delta$, and $t : \Tm\;\Delta\;A$, the term $[\sigma]t$ is propositionally equal to $[\sigma]_{\Tm}\,t$. \qed
\end{restatable}

Finally, we have to show that both $[\blank]_{\Ty}\blank$ and $[\blank]_{\Tm}\blank$ are coherent.
\begin{restatable}{proposition}{propCoherence}\label{prop:coherence-2}
  For every equality constructor $p : \sigma = \tau : \Sub\;\Gamma\;\Delta$, the identities
  \danger
  \[
    [\sigma]_{\Ty}\,A = [\tau]_{\Ty}\,A
    \quad\text{and}\quad
    [\sigma]_{\Tm}\,t = [\tau]_{\Tm}\,t
  \]
  hold for any type $A$ and any term $t$. \qed
\end{restatable}

@@

\begin{restatable}{proposition}{propConfluenceThree}\label{prop:local-confluence-3}
  Type substitution $[\blank]_{\Ty}\blank$ is terminating and locally confluent.
  \qed
    \danger
\end{restatable}

@@

\begin{restatable}{proposition}{propCorrectnessTwo}\label{prop:correctness-2}
    For every $\sigma : \Sub\;\Gamma\;\Delta$, and $A : \Ty\;\Delta$, the lifting
    $\sigma \uparrow A$ is propositionally equal to $\sigma^+$.
    \qed
\end{restatable}

@@

Finally, we show that these simultaneously defined functions are coherent.
\begin{restatable}{proposition}{propCoherenceUparrow} %[Coherence]
    For every equality constructor $p : \sigma = \tau : \Sub\;\Gamma\;\Delta$, we have
    \danger
    \[
      \sigma \uparrow A =^{\lam{\gamma}\Sub\;(\Gamma, [ \gamma ]_{\Ty} A)\;(\Delta, A)}_{\alert{p}} \tau \uparrow A
      \qquad\text{and}\qquad
      [\sigma]_{\Tm}\,t =^{\lam{\gamma}\Tm\;\Gamma\;([\gamma]_{\Ty}A)}_{\alert{p}} [\tau]_{\Tm}\,t,
    \]
    for any type $A$ and any term $t$.
    \qed
\end{restatable}

@@

Of course, we also have to prove coherence for the newly introduced recursive functions to verify that they are well defined.

\begin{restatable}%[{Coherence for $[\blank]_{\Tel}$ and $\blank\upuparrows\blank$}]
{proposition}{coherenceTel}
  For every equality constructor $p : \sigma = \tau$ and telescope $\Xi$, we have
  $[\sigma]_{\Tel}\,\Xi = [\tau]_{\Tel}\,\Xi$ and $\sigma \upuparrows \Xi =^{\Sub\;(\Gamma\dplus [\sigma]_{\Tel}\Xi)\;(\Delta\dplus\Xi)} \tau \upuparrows \Xi$. \qed
  \danger
\end{restatable}

@@

\begin{corollary}
  We have $[\sigma]_{\Ty}\,A = [\tau]_{\Ty}\,A$ for every equality constructor $p : \sigma = \tau$.
  \qed
  \danger
\end{corollary}

@@

The elimination principle computes by interpreting each constructor using the corresponding method inductively, e.g.,
\begin{alignat*}{5}
%  \elim_{\Ctx}\;\emptyctx   & \reduce \emptyctx_{\Ctx}                                    && : \Ctx^M\;\emptyctx  \\
  \elim_{\Ctx}\;(\Gamma, A) & \reduce (\elim_{\Ctx}\,\Gamma),_{\Ctx}^{M} (\elim_{\Ty}\,A) && : \Ctx^M\;(\Gamma, A) %\\
%  \elim_{\Ty}\;\UU          & \reduce \UU^M && :\Ty^M\;\Gamma^M\;\UU \\
%  \elim_{\Sub}\;\emptysub   & \reduce \emptysub^M_{\Sub} &&: \Sub^M\;\Gamma^M\;\emptyctx^M_{\Ctx}\;\emptyctx \\
%                            & \vdots
\end{alignat*}
%
Interestingly, a key difference compared to its QIIT counterpart is that $[\blank]_{\Ty}\blank$, $[\blank]_{\Tm}$, and $\blank\uparrow\blank$ are functions rather than (formed by) constructors, meaning they cannot be eliminated inductively but instead become propositional identities.
%type substitution $[\blank]\blank$ was a constructor and $[\sigma]\,A$ was eliminated to $[\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A)$; similarly for term substitution.
\begin{restatable}{proposition}{propProvableFunc}\label{prop:provable-function-clauses}
  Given the motives and the methods, we can prove
\begin{alignat*}{5}
  \elim_{\Ty}[] & : \elim_{\Ty}\,([\sigma]_{\Ty}\,A) && = && [\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A) \\
  \elim_{\Tm}[] & : \elim_{\Tm}\,([\sigma]_{\Tm}\,t) && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{\elim_{\Ty}[]}} && [\elim_{\Sub}\,\sigma]^M_{\Tm}\,(\elim_{\Tm}\,t) \\
  \elim_{\Sub}[] & : \elim_{\Sub}(\sigma\uparrow A) && =^{\lam{B^M}\Sub^M\;(\Gamma^M,^M B^M)\;(\Delta^M, A^M)\;(\sigma \uparrow A)}_{\alert{\elim_{\Ty}[]}} && (\elim_{\Sub} \sigma) \uparrow^M (\elim_{\Ty}\,A)
\end{alignat*}
\end{restatable}

@@

Going back and forth from one definition to the other gives us the equivalences
\begin{restatable}{theorem}{equivalence}
  Type theory defined as a QIIRT and a QIIT are equivalent, i.e.\ 
  \begin{itemize}
    \item $\eqv{\Ctx_{\IR}}{\Ctx_{\I}}$, 
    \item $\eqv{\left((\Gamma : \Ctx_{\IR})\to\Ty_{\IR}\;\Gamma\right)}{\left((\Gamma : \Ctx_{\I})\to\Ty_{\I}\;\Gamma\right)}$,
    \item $\eqv{\left((\Gamma\;\Delta : \Ctx_{\IR}) \to \Sub_{\IR}\;\Gamma\;\Delta\right)}%
               {\left((\Gamma\;\Delta: \Ctx_{\I}) \to \Sub_{\I}\;\Gamma\;\Delta\right)}$, and 
    \item $\eqv{\left((\Gamma : \Ctx_{\IR}) \to (A : \Ty_{\IR}\;\Gamma) \to \Tm_{\IR}\;\Gamma\;A\right)}
               {\left((\Gamma : \Ctx_{\I}) \to (A : \Ty_{\I}\;\Gamma) \to \Tm_{\I}\;\Gamma\;A\right)}$.
  \end{itemize}
\end{restatable}

@@

\section{Omitted proofs}\label{appendix:proofs}

\propConfluenceTwo*
\begin{proof}
  In addition to previous cases in the proof of \cref{prop:local-confluence-1}, consider terms 
  \[
    [\idS]_{\Ty}\;(\El\,u), \qquad [\sigma;\tau]_{\Ty}\;(\El\,u), \qquad [\pi_1(\sigma, t)]_{\Ty}\;(\El\,u), \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;(\El\,u)
  \]
  which reduce to the following terms
  \[
    \El\,u, \qquad \El([\sigma]_{\Tm}\,[\tau]_{\Tm}\,u), \qquad \El([\sigma]_{\Tm}\,u), \qquad\text{and}\qquad
    \El([\sigma]_{\Tm}\,[\pi_1\tau]_{\Tm}\,u)
  \]
  respectively, no matter which function clause of $[\blank]_{\Ty}\blank$ is used.
  Therefore, type substitution is locally confluent.
\end{proof}

\propCorrectness*
\begin{proof}
  We show this statement by induction on $\sigma$. Since we are proving a propositional equality, and we are assuming UIP, all equality constructors are automatically respected, so we only need to worry about the point constructors.
  \begin{enumerate}
    \item For the identity substitution $\idS$, we have $[\idS]t = t \judgeq [\idS]_{\Tm} t$ by definition.
    \item For a composite substitution $\sigma; \tau$, we have
      \begin{align*}
        [\sigma;\tau]t & = [\sigma]\,([\tau] t)            && \text{by $[;]t$} \\
%                       & = [\sigma]\,([\tau]_{\Tm}\,t)       && \text{by the induction hypothesis $[\tau]t = [\tau]_{\Tm}\,t$ for any $t$} \\
                       & = [\sigma]_{\Tm}\,([\tau]_{\Tm}\,t) && \text{by the induction hypothesis (twice)} \\ %$[\sigma]u = [\sigma]_{\Tm}\,u$ for any $u$} \\
                       & \judgeq [\sigma;\tau]_{\Tm}\,t && \text{by definition.}
      \end{align*}
    \item For $\pi_1(\sigma, t)$, we have
      \begin{align*}
        [\pi_1(\sigma, t)]t & = [\sigma] t               && \text{by $\pi_1\beta$} \\
                            & = [\sigma]_{\Tm} \,t       && \text{by the induction hypothesis} \\ % $[\tau]t = [\tau]_{\Tm}\,t$ for any $t$} \\
                            & \judgeq [\pi_1(\sigma, t)]_{\Tm}\,t && \text{by definition.}
      \end{align*}
    %\item For $\pi_1(\sigma; \tau)$, it is proved similarly as the previous case.
    \item For the remaining cases, $[\sigma]_{\Tm} t$ is equal to $[\sigma]\,t$ strictly.
  \end{enumerate}
  Hence, we conclude that $[\blank]_{\Tm}\blank$ is propositionally equal to $[\blank]\blank$.
\end{proof}
